Does at-issueness predict projection? Further investigations of the Gradient Projection Principle

### 2023-03-13 writing/SuB-abstract/SuB-abstract.tex
abstract edits sections 1+2

\begin{center}
	\textbf{\large%\thetitle
		Projection differs across embedding operators}
\end{center}

\vspace{-.7\baselineskip}
\noindent 
	We present experimental evidence \textbf{(i)} that the projection of the content of attitude complements differs between entailment-canceling operators, and \textbf{(ii)} that this by-operator variation differs between attitude predicates. {JT: I don't think point (i) is true when we aggregate across the predicates, no? And when we don't aggregate, point (i) is just point (ii). So perhaps we should just focus on point (ii)? I've written the theoretical discussion that way.} The observed variation is not captured by contemporary projection analyses (e.g., \citealp{heim83,vds92,abrusan2011,schlenker2021}) and does not align with the long-standing distinction of factive vs. semi-factive predicates (e.g., \citealp{karttunen_observations_1971,hooper_applicability_1973,djarv_cognitive_2018}). Rather, the observed by-operator variation reveals a more nuanced picture of lexical semantic/pragmatic properties which raises important questions for future research on projection.

\noindent
	{\bf Projection across entailment-cancelling operators.} Interpreters may infer that a speaker who utters an attitude ascription, as in \ref{ex:family}, is committed to the content of the complement (CC),
	even when it occurs under an entailment-canceling operator, like negation \ref{ex:neg}, polar questions \ref{ex:q}, modals \ref{ex:mod}, or conditionals \ref{ex:cond}.
	\vspace{-.6\baselineskip}
	\ex. \label{ex:family}
		\a. \label{ex:neg}
			{\bf Negation:} \hfill
			\emph{\lq Cole didn't discover that Julian dances salsa.\rq}
		\b. \label{ex:q}
			{\bf Polar Question:} \hfill
			\emph{\lq Did Cole discover that Julian dances salsa?\rq}
		\c. \label{ex:mod}
			{\bf Modal:} \hfill
			\emph{\lq Perhaps Cole discovered that Julian dances salsa.\rq}
		\d. \label{ex:cond}
			{\bf Conditional:} \hfill
			\emph{\lq If Cole discovered that Julian dances salsa, Logan will be joyful.\rq}
		\z.
	\z.

\vspace{-.6\baselineskip}
Contemporary research rarely considers potential differences in projection from under these four operators, with very few exceptions and conflicting results:  
\begin{itemize}[topsep = 0pt, itemsep=-5pt, leftmargin = 12pt]
\item \citet{smith_relationship_2014} observed that the content of non-restrictive relative clauses and the preparatory content of \emph{win} was more projective under conditionals than negation, while the content of the clausal complement (CC) of English \emph{know} and the projective content of epithets exhibited the opposite pattern. In contrast, \citet{siekers-solstad2022} did not observe that the CC of German \emph{wissen} `know' was more projective under conditionals than negation. 
\item \citet{karttunen_observations_1971} suggested that the CC of English factive predicates (e.g. \emph{be annoyed, regret}) projects across all four operators, whereas that of English semi-factive predicates (e.g. \emph{discover, realize, see, notice}) always projects from under negation, but not always from the other three operators. This suggestion does not find empirical support from the experiment reported on in \citealt{siekers-solstad2022} for German, where the CC of the factive predicates \emph{bereuen} `regret', \emph{wissen} `know', and \emph{enth\"ullen} `reveal' did not project more from under negation than that of the semi-factive predicates \emph{entdecken} `discover', \emph{bemerken} `notice', and {\em herausfinden} `find out'. 
\end{itemize}

\noindent
To investigate whether these diverging results are due to differences between English and German or differences in the methods of investigation, we ran a series of experiments designed to investigate projection from under the four entailment-canceling operators in \ref{ex:family}. Our experiment used the same projection measure as \citealt{siekers-solstad2022} (the `certain that' diagnostic; see e.g., \citealp{tonhauser_how_2018,djarv_prosodic_2017,mahler_social_2020}) and applied this diagnostic to the contents of the complements of 20 English clause-embedding predicates, including purported factive predicates (e.g., \emph{be annoyed, know, reveal}) and purported semi-factive predicates (e.g., \emph{discover, see}). Given recent results that the CC of non-factive predicates is also projective (\citealt{degen-tonhauser-language}), albeit to varying degrees, we also included 15 non-factive predicates.

@@

@@

{\bf THIS IS OLD, FROM ABOVE: So far, there has been no experimental investigation of by-operator variation comparing factive and semi-factive predicates. However, \citet{djarv_cognitive_2018} and \citet{tonhauser_how_2018} observed by-predicate variation in polar questions. \citet{djarv_cognitive_2018}, assessing acceptability of affirming the main clause while denying the CC, found higher ratings for \emph{be happy} and \emph{appreciate} (assumed to be factive) and \emph{be aware} than \emph{realize} (assumed to be semi-factive). Here, it is also not obvious how exactly this task relates to projection. \citet{tonhauser_how_2018} measured projection of the CC of a broad range of attitude predicates more directly, collecting ratings about speaker certainty about the CC. The observed differences did not match the expectations from Karttunen's classification (e.g., the CC of semi-factive \emph{realize} was as projective as that of factive \emph{be annoyed} and more than that of semi-factive \emph{discover}).}

@@

\bibliography{../bibliography.bib}

>>>>

\begin{center}
	\textbf{\large%\thetitle
		Projection variability across embedding operators and triggers}
\end{center}

\vspace{-.7\baselineskip}
\noindent 
	We present experimental data which suggests that \textbf{(i)} the projectivity of the content of attitude complements varies depending on the entailment-canceling operator used, and \textbf{(ii)} that this effect of entailment-cancelling operator between attitude predicates.
	%
	Our finding of (ii) has not been previously observed and presents new theoretical challenges. The variability we found is gradient, confirming previous research that by-predicate variability is not determined by categorical lexical classes (\citealt{tonhauser_how_2018,degen_are_2022}), extending this result to by-predicate variability in interaction with embedding contexts.
	%
	Existing theoretical accounts of projection across different entailment-canceling operators cannot explain the observed by-trigger differences (e.g.,
	\citealt{heim_projection_1983,van_der_sandt_presupposition_1992,abrusan_predicting_2011,schlenker_triggering_2021}).
	% \citealp{heim83,vds92,abrusan2011,schlenker2021}).
	To our knowledge, the only approach that considers interactions between predicates and embedding contexts is the traditional classification of factive versus semi-factive predicates (e.g., \citealt{karttunen_observations_1971,hooper_applicability_1973,djarv_cognitive_2018}).
	%
	Our experimental data suggests that any account of projection must consider interactions between triggers and embedding operators, and that this interaction cannot be fully determined based on categorical lexical classes. Instead, the projection variability reveals a more nuanced picture of lexical semantic/pragmatic properties which raises important questions for future research on projection.

\noindent
	{\bf Investigating Projectivity across entailment-cancelling operators.} Language users may infer that a speaker who utters an attitude ascription, as in \ref{ex:family}, is committed to the content of the complement (CC),
	even when it occurs under an entailment-canceling operator, like negation \ref{ex:neg}, polar questions \ref{ex:q}, modals \ref{ex:mod}, or conditionals \ref{ex:cond}, in which case we say that it \emph{projects}.
	% \vspace{-.6\baselineskip}
	\ex. \label{ex:family}
		\a. \label{ex:neg}
			{\bf Negation:} \hfill
			\emph{\lq Cole didn't discover that Julian dances salsa.\rq}
		\b. \label{ex:q}
			{\bf Polar Question:} \hfill
			\emph{\lq Did Cole discover that Julian dances salsa?\rq}
		\c. \label{ex:mod}
			{\bf Modal:} \hfill
			\emph{\lq Perhaps Cole discovered that Julian dances salsa.\rq}
		\d. \label{ex:cond}
			{\bf Conditional:} \hfill
			\emph{\lq If Cole discovered that Julian dances salsa, Logan will be joyful.\rq}
		\z.
	\z.

	\noindent Following the methodology and approach from \citet{tonhauser_how_2018,degen_are_2022}, who find gradient differences in projection among different clause-embedding predicates, we talk about \emph{projectivity} as the degree or frequency with which a projective interpretation is assigned. 
%
	Contemporary empirical research rarely considers potential differences in projectivity from under these four operators, with very few exceptions and conflicting results: \citet{smith_relationship_2014} investigated by-operator variation (negation, conditional) for various types of projective content, observing that the content of non-restrictive relative clauses and the preparatory content of \emph{win} was more projective under conditionals than negation, while the projective content of epithets and the clausal complement (CC) of English \emph{know} exhibited the opposite pattern. In contrast, \citet{sieker_projective_2022}, investigating projective variabiltiy of the CC German attitude predicates across the four operators in \Last, they found no significant projection variability across operators, and did not report interactions by-predicate. In particular they did not find a difference between factive and semi-factive verbs, nor did they replicate \posscite{smith_relationship_2014} result for the German counterpart of `know' (\emph{wissen}). This also raises the question whether these diverging results are due to differences between English and German or differences in the methods of investigation.
	%

	To investigate projection variability across operators and attitude predicates, we ran a series of experiments designed to investigate projection from under the four entailment-canceling operators in \ref{ex:family}. Our experiment used the same projection measure as \citealt{sieker_projective_2022} (the `certain that' diagnostic; see e.g., \citealp{tonhauser_how_2018,djarv_prosodic_2017,mahler_social_2020}) and applied this diagnostic to the contents of the complements of 20 English clause-embedding predicates, including purported factive predicates (e.g., \emph{be annoyed, know, reveal}) and purported semi-factive predicates (e.g., \emph{discover, see}). Given recent results that the CC of non-factive predicates is also projective (\citealt{degen_are_2022}), albeit to varying degrees, we also included 15 non-factive predicates.

@@

\begin{itemize}
	\item \citet{karttunen_observations_1971} suggested that the CC of English factive predicates (e.g. \emph{be annoyed, regret}) projects across all four operators, whereas that of English semi-factive predicates (e.g. \emph{discover, realize, see, notice}) always projects from under negation, but not always from the other three operators. This suggestion does not find empirical support from the experiment reported on in \citealt{siekers-solstad2022} for German, where the CC of the factive predicates \emph{bereuen} `regret', \emph{wissen} `know', and \emph{enth\"ullen} `reveal' did not project more from under negation than that of the semi-factive predicates \emph{entdecken} `discover', \emph{bemerken} `notice', and {\em herausfinden} `find out'. 
\end{itemize}

@@

% {\bf THIS IS OLD, FROM ABOVE: So far, there has been no experimental investigation of by-operator variation comparing factive and semi-factive predicates. However, \citet{djarv_cognitive_2018} and \citet{tonhauser_how_2018} observed by-predicate variation in polar questions. \citet{djarv_cognitive_2018}, assessing acceptability of affirming the main clause while denying the CC, found higher ratings for \emph{be happy} and \emph{appreciate} (assumed to be factive) and \emph{be aware} than \emph{realize} (assumed to be semi-factive). Here, it is also not obvious how exactly this task relates to projection. \citet{tonhauser_how_2018} measured projection of the CC of a broad range of attitude predicates more directly, collecting ratings about speaker certainty about the CC. The observed differences did not match the expectations from Karttunen's classification (e.g., the CC of semi-factive \emph{realize} was as projective as that of factive \emph{be annoyed} and more than that of semi-factive \emph{discover}).}

@@

% \bibliography{../bibliography.bib}
\bibliography{../projective-content.bib}
